---
title: "Homework 2"
output:   
  github_document
---

### Load packages
```{r, message = FALSE}
library(tidyverse)
library(readxl)
```

## Problem 1

### Import NYC Transit data and perform initial cleaning
```{r}
transit_df <- read_csv('data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv',
                       col_types = cols(Route8 = "c", Route9 = "c", Route10 = "c", Route11 = "c")) %>%
  janitor::clean_names() %>%
  select(line:entry, vending, ada) %>%
  mutate(entry = ifelse(entry == "YES", TRUE, FALSE))
```

*`transit_df` includes variation NYC subway information by station. Information included includes the subway line that runs through the station, the station name, latitude and longitude of the station, various routes served by said station, the type of entrance (stair, door, easement, elevator, escalator, ramp, walkway), if entrance is allowed, if station has vending, and if the station is ADA compliant.* 

*We adjusted `Route8`, `Route9`, `Route10`, and `Route11` to character values. `janitor::clean_names()` adjusted variable names to snake case. We then selected key variables and adjusted `entry` to a logical variable. At this stage, the dataset is not yet "tidy" as the route variables should be in long format.*


###### How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.
```{r}
transit_df %>% 
  select(station_name, line) %>%
  distinct() %>%
  summarize(n = n())
```
*There are a total of `465` unique stations in the NYC subway.*

###### How many stations are ADA compliant?
```{r}
transit_df %>%
  filter(ada == TRUE) %>%
  select(station_name, line) %>%
  distinct() %>%
  summarize(n = n())
```
*`84` unique stations are ADA compliant.*

###### What proportion of station entrances / exits without vending allow entrance?
```{r}
transit_df %>%
  filter(vending == "NO") %>%
  group_by(entry) %>%
  summarize(n = n()) %>%
  mutate(prop = n/sum(n))
```
*Of station entrances/exits without vending, `37.7%` allow entrance.*

###### Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?
```{r}
transit_df %>%
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route"
  ) %>%
  filter(route == "A") %>%
  select(station_name, line) %>%
  distinct() %>%
  summarize(n = n())


transit_df %>%
  pivot_longer(
    route1:route11,
    names_to = "route_num",
    values_to = "route"
  ) %>%
  filter(route == "A", ada == TRUE) %>%
  select(station_name, line) %>%
  distinct() %>%
  summarize(n = n())
```

*`60` stations serve the A train. Of stations that serve the A train, `17` stations are ADA compliant.*

## Problem 2

### Import Trash Wheel dataset and perform cleaning
```{r}
mr_trash_df <- read_excel('data/Trash Wheel Collection Data.xlsx',
                       sheet = "Mr. Trash Wheel",
                       range = "A2:N549",
                       col_names = TRUE) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(sports_balls = as.integer(round(sports_balls)),
         dataset_name = "Mr. Trash Wheel",
         date = str_replace(as.character(date), "1900-01-20", "2020-01-20"),
         date = as.POSIXct(date))

prof_trash_df <- read_excel('data/Trash Wheel Collection Data.xlsx',
                       sheet = "Professor Trash Wheel",
                       range = "A2:M96",
                       col_names = TRUE) %>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(dataset_name = "Professor Trash Wheel",
         year = as.character(year))  

trash_df <- bind_rows(mr_trash_df, prof_trash_df) %>%
  arrange(date) %>%
  mutate(date = as.Date(date))
```

*`trash_df` contains information about trash collection efforts by 2 water-wheel vessel that removes trash from the Inner Harbor in Baltimore, Maryland. Trash collected is sent to an energy plant, converting it into electricity. There are a total of `r nrow(trash_df)` observations after merging `mr_trash_df` and `prof_trash_df`. Key variables include `date`, `weight_tons`, `volume_cubic_yards`, `plastic_bottles`, and `homes_powered`. Total weight of trash collected by Professor Trash Wheel was `r sum(prof_trash_df$weight_tons)`. The total number of sports balls collected by Mr. Trash Wheel in 2020 was `r mr_trash_df %>% filter(year == 2020) %>% summarize(sum(sports_balls))`.*

## Problem 3

### Import FiveThirtyEight data and perform cleaning
```{r, collapse = TRUE}
pols <- read_csv('data/pols-month.csv',
                 col_names = TRUE,
                 show_col_types = FALSE) %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(month = month.name[as.numeric(month)],
         president = case_when(
           prez_gop == 1 ~ "gop",
           prez_dem == 1 ~ "dem",
           TRUE ~ "NA"
         )) %>%
  select(-c(prez_dem, prez_gop, day)) %>%
  arrange(year, month)

snp <- read_csv('data/snp.csv',
                col_names = TRUE,
                show_col_types = FALSE) %>%
  janitor::clean_names() %>%
  separate(date, into = c("month", "day", "year"), sep = "/") %>%
  mutate(month = month.name[as.numeric(month)],
         year = ifelse(as.numeric(year) <= 99 & as.numeric(year) > 15, as.numeric(year) + 1900, as.numeric(year) + 2000),
         year = as.character(year)) %>%
  relocate(year, .before = month) %>% 
  select(-day) %>%
  arrange(year, month)

ue <- read_csv('data/unemployment.csv',
               col_names = TRUE,
               show_col_types = FALSE) %>%
  pivot_longer(
    Jan:Dec,
    names_to = "month",
    values_to = "percent_unemployed"
  ) %>%
  rename(year = Year) %>%
  mutate(year = as.character(year),
         month = month.name[match(month, month.abb)]) %>%
  arrange(year, month)
```

### Join datasets (merge `snp` with `pols`, then merging `ue` with result)
```{r}
snp_pols <- pols %>% 
  left_join(snp, by = c("year", "month"))

final_df <- snp_pols %>%
  left_join(ue, by = c("year", "month"))
```

*`pols` contains data on political affiliation of the sitting president and the number of democratic/republican governor/senators/representatives on the by month and year. `snp` contains information the closing values of the  S&P index by date. `ue` includes unemployment data in percentage of unemployment by month and year.*

*After merging all three datasets, the resulting dataset has a total of `r nrow(final_df)` observations and `r ncol(final_df)` variables. Year of data ranges between `r min(final_df$year)` and `r max(final_df$year)`. Key variables include `year`, `month`, `president`, `close`, and `percent_unemployed`.* 

